Semi-quantitative petrographic characterization of Bronze Age quartzite tools – an exploratory approach

### 2022-12-07 analysis/paper/paper.Rmd
Updated discussion with reference to graphite-schists
Continued update and rewrite of manuscript and layout

Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF). The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. Petrographic characterization of the material represents the important first step in the reconstruction of the *chaîne opératoire*, as well as questions regarding raw material selection and management. Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions] has been highlighted by Prieto et al. [-@prieto_etal19 p. 15]. In Northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region. While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden has been put forth [@halen94 pp. 115-122; @lundberg97 pp. 161-171], there has been little development on the topic of quartz and quartzite. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape. The tentative interpretations of lithic material sourcing within Northern Sweden tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05]. Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies. In the case of archaeological research it is worth considering situations where destructive techniques simply are not applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information. Aside from being non-destructive spectroscopic instrumentations feature fast measurements, making it an attractive option for field-based screening purposes. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].

In the current study a dataset of quartz and quartzite from Vilhelmina and Åsele parishes in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF). The study includes points and preforms collected from `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (fig. \@ref(fig:nir-spectra-summary)). The aim is to assess the potential of non-destructive techniques for characterization of quartz and quartzite material without the use of references from known geological sources. Due to the geological complexity of both the region, as well as the material, raw material management studies struggles with a significant amount of uncertainties. This study represents a first look into the chemical characteristic of the material, and will test whether the field-based instrumentation can be used to identify the variation between the samples in  order to form the basis for a predictive model.

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden. The sites are distributed along the Ångermanälven river valley, from the lake Kultsjön in the north, down towards the town of Åsele in the south. A dataset of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF. The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint, with some being almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent).

A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, @mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89], suggesting a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].

@@

The dataset includes three main groups of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe. "*brecciekvarts*"). Those that had no prior classification available were sorted into either the quartz or quartzite category based on visual inspection. This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples. The inclusion of quartz and brecciated quartz in the study serves two purposes. The first being to test the potential for separating quartz from quartzite based on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.

@@

All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/en/research/infrastructure/mal/]. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may be of the same material, however, for the smaller sites there still remains some uncertainties. For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.

Data analysis and visualisation was performed in R [@Rteam21] using a suite of packages, including Chemospec [@hanson16], missMDA [@josse_husson16], ggplot2 [@wickham16] and plotly [@sievert20]. Pre-processing and analysis of the data is different for each instrumentation and based on relevant literature. Data exploration and analysis was performed using Principal Component Analysis (PCA).

@@

### Raman
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01]. However, where NIRS measures light absorbance and generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

@@

After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data. This was achieved using modified polynomial fitting in R [@Rteam21] with the ChemoSpec package [@hanson16]. The spectra were then mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] before PCA.

@@

### X-Ray Fluourescence
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method for petrographic analysis in archaeology [@hall60; @shackley11a]. Exposing a material to X-ray radiation causes electrons to dislodge and move between the electron shells of the atoms, which in turn emits energy (i.e. *fluourescence*). As the energy differences between electron shells for the various elements is known it is possible to both identify and quantify the abundance of an element [@shackley11b p.16], which is useful for provenience studies where the chemical character of an unknown and a reference sample can tested statistically [@tykot03].

XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was set to *mining* mode [@knight_etal21 p.2]. The integration time was set to 120 seconds in order for the instrument to cycle through all of the filters (main, low, high, light). Due to time constraints, and the number of objects included in the project, it was only possible to sample each object once, apart from some exceptions.

The XRF data was transformed using *centred log-ratio transformation*, mean-centered and scaled to unit variance before PCA. As the resulting dataset also features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.

Multivariate analysis methods, such as PCA, are also useful for overcoming some of the issues related to so-called "data closure" [@reimann_etal12]. The compositional data add up to a constant (100%), which is described as the *closed sum* effect. Correlation analysis is directly affected by this *closure* effect and therefore not suitable. The application of *centred log-ratio transformation* was done in order to address the closed sum effect, and in combination PCA it can reveal the structure and relations between the variables (elements) more accurately.

@@

The data was evaluated using PCA, which is an unsupervised exploratory technique to reduce the dimensionality of a dataset [@geladi_linderholm20]. PCA has proven useful for exploring complex spectroscopic datasets [@sciuto18], and identifying clusters and trends that relate to the spectral fingerprints of the material. These groupings can then be followed up with supervised methods, and in the construction of classification models.

@@

#Filter out the dark quartzite samples that feature peaks of interest in their spectra 
points.d.filt <-
  metadata.csv %>%
  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
           site_id == "Åsele 99", 
         type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
    filter(!sample_id %in% c("153","167","168","169","172","174","175","177","182","183","190","191","193","194","196","198","200","204","207","210","213","214",
                           "215","216","229","234","235","237","238","251","262","265","268","269","272","278","281","282","359","377","385","392","393","397","405",
                           "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56")) %>% 
  filter(hue == "Dark")

@@

It is well-known that NIRS is less effective on translucent material, such as a pure quartz crystal, as the absence of impurities in the material will result in little to no reflection/absorbance of light [@hunt77]. Similarly it can be difficult to work with dark materials as most of the light produced by the instrument will be absorbed, with the high absorbance potentially obscuring visible features in the spectra. This effect can be seen in figure \@ref(fig:nir-spectra-summary) which shows a summary of the centered NIR spectra by color group. The dark group generally features less distinct peaks than the material with a lighter hue. This is not to say that all spectra collected on dark quartzite features less information, but any model including them will likely affect the clustering of the dataset as a whole.

The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17 pp.78-81]. 2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the cost of potentially amplifying noise [@mark_workman21 pp. 351-397; @schmid_etal22]. An example of this can be seen in fig. \@ref(fig:nir-savitzky-transform), where one of the centered spectra from the dark group is compared to its derivative counterpart. The SG transformation amplifies a number of significant peaks in the spectra, as well as a shift in the instrumentation filter at 1800 nm. Out of a total of `r length(spec.d.total)` dark spectra `r length(spec.d.filtered)` were kept in the dark group after filtering out those spectra that feature no discernible peaks or mainly noise. Despite this there is still a considerable difference in light absorbance between the dark and light group, as can be seen by comparing fig. \@ref(fig:nir-spectra-summary) and fig. \@ref(fig:nir-savitzky-transform):A.

@@

```{r nir-spectra-summary, fig.cap= "nir spectra summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/001-nir-spectra-summary.R"), echo = FALSE}
quietly(source("./analysis/paper/001-nir-spectra-summary.R", encoding = "UTF-8"))
knitr::include_graphics('../figures/001-nir-spectra-summary.png', error = FALSE)
```

@@

```{r nir-savitzky-transform, fig.cap= "nir savitzky comparison", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform.R"), echo = FALSE}
quietly(source("./analysis/paper/003-nir-savitzky-transform.R", encoding = "UTF-8"))
knitr::include_graphics("../figures/003-nir-savitzky-transform.png", error = FALSE)
```

@@

Based on the evaluation of the spectra a couple of significant peaks can be identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (fig. \@ref(fig:nir-savitzky-transform-alt) & \@ref(fig:nir-peak-summary)) [@hunt77 p. 508-510; @sciuto_etal19]. OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region. As a general rule, the 2 200 nm and 2 350 nm peak positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77 p.509; @clark_etal90 p.12664]. While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH. In the current spectra the 2 200 nm band is typically the more intense band.

The above absorption bands can be found in minerals such as muscovite, which would not be strange to find in quartzite [@clark_etal90 p.12; hunt77 p.510; @ramanaidou15 p.207]. Although the 2 350 nm band is one of the stronger carbonate (CO) bands these are typically not associated to water features, and considering that the 2 350 nm peak always is accompanied by the 2 200 nm peak it would be unlikely to be related to CO stretching in this case.

>>>>

Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF). The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. 
Characterization of the material, and identification of its potential source(s), represents the important first step in the reconstruction of the *chaîne opératoire*, and inferences of strategies for raw material selection and management. Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these materials. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions] has been highlighted by Prieto et al. [-@prieto_etal19 p. 15]. In Northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region as they represent the most common material groups at archaeological sites. While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden has been put forth [@halen94 pp. 115-122; @lundberg97 pp. 161-171], in contrast, there has been little development on the topic of quartz and quartzite. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape. What interpretations of lithic material sourcing within Northern Sweden there are tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05]. Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies. Spectroscopic instrumentations feature fast measurements that can be collected directly on the surface of an object. With newer mobile models of these instruments readily available, this makes it an attractive option for field-based screening purposes. In the case of archaeological research it is also worth considering situations where destructive techniques, such as thin-section analysis, are not suitable or applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other complementary instrumentations that may provide further insight into the structural properties of a material. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. The same can be said for Raman spectroscopy, which has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, and the Earth sciences [@dubessy_etal12].

In the current study a dataset of quartz and quartzite from Vilhelmina and Åsele parishes in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF). The study includes points and preforms collected from `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (fig. \@ref(fig:nir-spectra-summary)). The aim is to assess the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources. The geological complexity of both the region, as well as the material, necessitates alternative approaches to studies aimed at raw material management. Without relying on well-delimited source areas for geological material, a predictive model capable of separating different "classes" of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models. The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and thus this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden during the 4 000 - 2 000 BP period.

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden. The sites are distributed along the Ångermanälven river valley, from the lake Kultsjön in the north, down towards the town of Åsele in the south. A dataset of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF. The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint, with some being almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent).

A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, @mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89], suggesting a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite when compared to prior periods [@baudou78].

@@

The dataset includes three main groups of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe. "*brecciekvarts*"). Those that had no prior classification available were sorted into either the quartz or quartzite category based on visual inspection. This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples. The inclusion of quartz and brecciated quartz in the study serves two purposes. The first being to test the potential for the instrumentation to separate quartz from quartzite based on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.

@@

All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/en/research/infrastructure/mal/]. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult. For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix. An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may come from the same geological source, but the same can not be said for sites with smaller assemblages.

Data analysis and visualisation was performed in R [@Rteam21] using a suite of packages, including Chemospec [@hanson16] and zCompositions [@palarea-albaladejo_martin-fernandez15] for data processing, and ggplot2 [@wickham16] for visualisation. Pre-processing and analysis of the data is different for each instrumentation and based on relevant literature. Data exploration and analysis was performed using Principal Component Analysis (PCA).

@@

### Raman
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared spectrum [@nafie01]. However, where NIRS measures light absorbance and generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

@@

After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data. This was achieved using modified polynomial fitting [@lieber_mahadevan-jansen03] from the ChemoSpec package [@hanson16]. The spectra were then mean-centred <!--and normalized using Standard Normal Variate [@shaver01 pp.284-285] --> before PCA.

@@

### X-Ray Fluourescence
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Exposing a material to X-ray radiation causes electrons to dislodge and move between the electron shells of the atoms, which in turn emits energy (i.e. *fluourescence*). As the energy differences between electron shells for the various elements is known it is possible to both identify and quantify the abundance of an element [@shackley11b p.16], which is useful for provenience studies where the chemical character of an unknown and a reference sample can be tested statistically [@tykot03].

XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was set to *mining* mode [@knight_etal21 p.2]. The integration time was set to 130 seconds in order for the instrument to cycle through all of the filters (main, low, high, light). Due to time constraints, and the number of objects included in the project, it was only possible to sample each object once, apart from some exceptions.

Before engagaging in any analysis it is important to address the *closed sum* effect of compositional data [@pawlowsky-glahn_egozcue06]. As the variables are given as proportions of a constant sum (e.g. percentage, parts per million) this means that if one element was to change in value, others would follow [@grunsky_caritat20]. This is managed by transforming the data using log-ratio transformation [@aitchison99], and currently there are three main methods for achieving this; additive log-ratio (alr), centered log-ratio (clr) or isometric log-ratio. They all have their own nuances and as such the choice of method is dependent on the type of analysis and material [@pawlowsky-glahn_etal15]. In the PCA the Principal Components are orthonormal, and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g. @reimann_etal12]. This makes multivariate methods like PCA suitable for identifying geochemical trends related to the rock-forming process.

As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15]. *lrEM* stands for "Log-ratio Expectation-Maximisation", which is an algorithm used to impute "left-censored data" (e.g. values below LOD, rounded zeros). Following log-transformation the data was then mean-centered and scaled to unit variance before PCA.

@@

The data was evaluated using PCA, which is an unsupervised exploratory technique to reduce the dimensionality of a dataset [@geladi_linderholm20]. Multivariate approaches, such as PCA,  has proven useful for exploring complex spectroscopic datasets [@sciuto18]. Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material. These groupings can then be followed up with supervised methods, and aid in the construction of predicitive models.

@@

#Filter out the dark quartzite samples that feature peaks of interest in their spectra 
points.d.filt <-
  metadata.csv %>%
  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
           site_id == "Åsele 99", 
         type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
  filter(!sample_id %in% c("153","167","168","169","172","174","175","176","177","182","190","191","193","194","196","198","200","204","214",
                           "215","216","229","238","262","265","272","282","359","391","392","393","397","405",
                           "408","410","411","413","414","415","416","417","424","425","428","430","432","55","56")) %>% 
  filter(hue == "Dark")

@@

It is well-known that NIRS is less effective on translucent material, such as a pure quartz crystal, as the absence of impurities in the material will result in little to no reflection/absorbance of light [@hunt77]. Similarly it can be difficult to work with dark materials as most of the light produced by the instrument will be absorbed, with the high absorbance potentially obscuring visible features in the spectra. This effect can be seen in figure \@ref(fig:nir-spectra-summary) which shows a summary of the centered NIR spectra by color group. Apart from the dark material featuring a wider range of absorbance, likely owing some to the range of hues within the group, they generally features less distinct peaks than the material with a lighter hue. This is not to say that all spectra collected on dark quartzite features less information, but any model including them will likely affect the clustering of the dataset as a whole.

The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17 pp.78-81]. 2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the cost of potentially amplifying noise [@mark_workman21 pp. 351-397; @schmid_etal22]. An example of this can be seen in fig. \@ref(fig:nir-savitzky-transform), where the spectra from two different dark objects are compared. The non-processed spectra are fairly indistinct, but the derivative transformation of sample 258 amplifies a number of significant peaks. The derivative spectra of sample 411, however, does not have as clear and distinct peaks and features mainly noise. In both cases an interesting feature at 1 800 nm have also been amplified, which happens to be a filter shift in the instrumentation. After reviewing the dark spectra, out of a total of `r length(spec.d.total)` spectra, `r length(spec.d.filtered)` were kept in the dark group.

@@

```{r nir-spectra-summary, fig.cap= "Summary of the raw spectra for each colour group. Note the differeces in absorbance between the groups, with the dark group featuring featuring a larger range of absorbance values as well as less distinct peaks.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/001-nir-spectra-summary.R"), echo = FALSE}
quietly(source("./analysis/paper/001-nir-spectra-summary.R", encoding = "UTF-8"))
knitr::include_graphics('../figures/001-nir-spectra-summary.png', error = FALSE)
```

@@

```{r nir-savitzky-transform, fig.cap= "Comparison of NIR spectra between two samples from the dark group. a) the non-processed spectra for sample 258 and 411, b) sample 258 spectra with 2^nd^ derivative Savitzky-Golay filtering, c) sample 411 spectra with 2^nd^ derivative Savitzky-Golay filtering.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform2.R"), echo = FALSE}
quietly(source("./analysis/paper/003-nir-savitzky-transform2.R", encoding = "UTF-8"))
knitr::include_graphics("../figures/003-nir-savitzky-transform.png", error = FALSE)
```

@@

Based on the evaluation of the spectra a couple of significant peaks can be identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (fig. \@ref(fig:nir-savitzky-transform-alt) & \@ref(fig:nir-peak-summary)) [@hunt77 p. 508-510; @sciuto_etal19]. OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region. As a general rule, the 2 200 nm and 2 350 nm positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77 p.509; @clark_etal90 p.12664]. While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH. In the current dataset the 2 200 nm band is typically the more intense band, although there are exceptions.

The above absorption bands can be found in mica, such as muscovite and biotite, which would not be strange to find in quartzite [@clark_etal90 p.12; @hunt77 p.510; @ramanaidou_etal15 p.207]. Although muscovite and biotite share some band positions, the altered Al content as a response to metamorphism has been documented to shift some positions and increase others, such as the 2250 nm band for biotite [@duke94]. As can be seen in the average spectra in fig. \@ref(fig:nir-peak-summary) the dark material exhibits an additional peak at ~2 250 nm, which may indicate the presence of biotite and a different metamorphic grade.
